---
title: "Blue Bikes EDA"
author: "Justin Wells"
output: html_document
date: "2024-12-04"
---

## **still needs work 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
setwd("/Users/jmwells24/Desktop 2/DATA-0200/Final Project")
```
```{r}
# Load necessary libraries
library(dplyr)
library(readr)
library(ggplot2)
```
```{r}
# Define custom colors for MBTA lines
mbta_colors <- c(
  "BLUE" = "blue",
  "BLUE/GREEN" = "cyan",
  "BLUE/ORANGE" = "coral", 
  "GREEN" = "darkgreen", 
  "GREEN/ORANGE" = "brown",
  "GREEN/RED" = "green",
  "ORANGE" = "orange",
  "RED" = "red",
  "SILVER" = "gray"
)
```
```{r}
data <- read.csv("trip_data_with_mbta.csv")
```

# Dataset Overview
```{r}
# Structure of dataset
str(data)

# Summarize the dataset
summary(data)
```
# Dataset View
```{r}
# view columns
colnames(data)

# view first few rows
head(data)
```
# Data Inaccuracies
```{r}
# Check for null values
cat("Null values in each column:\n")
print(colSums(is.na(data)))

# Remove rows with null values
data <- na.omit(data)

# Verify if null values are removed
cat("Are there any null values left? ", any(is.na(data)), "\n")
```
#Descriptive Statistics
```{r}
aggregate(trip_duration ~ member_casual, data, mean)
```
#Proximity Analysis
```{r}
# Start Station Proximity
table(data$member_casual, data$start_station_proximity)
```
```{r}
# End Station Proximity
table(data$member_casual, data$end_station_proximity)
```
# Member vs. Casual Trends → Trip Duration
*Analyze and compare the average trip duration for members and casual riders*
```{r}
# Summarize trip duration for member and casual riders
member_casual_summary <- data %>%
  group_by(member_casual) %>%
  summarise(
    mean_trip_duration = mean(trip_duration, na.rm = TRUE),
    median_trip_duration = median(trip_duration, na.rm = TRUE),
    trip_count = n()
  )
print(member_casual_summary)
```
*Explore trip duration distributions for each rider type.*
```{r}
# Trip duration distribution by rider type
member_casual_distribution <- data %>%
  group_by(member_casual) %>%
  summarise(
    min_duration = min(trip_duration, na.rm = TRUE),
    max_duration = max(trip_duration, na.rm = TRUE),
    sd_duration = sd(trip_duration, na.rm = TRUE)
  )
print(member_casual_distribution)
```
# Time Trends → Trip Duration by Hour
*Breakdown of hourly trends*
```{r}
# Extract hour from the timestamp
data$hour <- format(as.POSIXct(data$started_at), "%H")

# Summarize trip duration by hour
time_trends <- data %>%
  group_by(hour) %>%
  summarise(
    mean_trip_duration = mean(trip_duration, na.rm = TRUE),
    median_trip_duration = median(trip_duration, na.rm = TRUE),
    trip_count = n()
  )
print(time_trends)
```
*Break down hourly trends by rider type (member vs. casual).*
```{r}
# Hourly trends by rider type
hourly_member_casual <- data %>%
  group_by(hour, member_casual) %>%
  summarise(
    mean_trip_duration = mean(trip_duration, na.rm = TRUE),
    trip_count = n()
  )
print(hourly_member_casual)
```

*Analyze how trip duration changes throughout the week (weekday vs. weekend).*
```{r}
# Weekday vs Weekend trends
data$day_type <- ifelse(weekdays(as.Date(data$started_at)) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
weekday_weekend_trends <- data %>%
  group_by(day_type) %>%
  summarise(
    mean_trip_duration = mean(trip_duration, na.rm = TRUE),
    trip_count = n()
  )
print(weekday_weekend_trends)

# Plot: Weekday vs Weekend Trends in Trip Duration
ggplot(weekday_weekend_trends, aes(x = day_type, y = mean_trip_duration, fill = day_type)) +
  geom_bar(stat = "identity") +
  ggtitle("Weekday vs Weekend Trends in Trip Duration") +
  xlab("Day Type") +
  ylab("Average Trip Duration (minutes)") +
  theme_minimal()
```
*Analyze how trip duration changes throughout the week (weekday vs. weekend) by rider type.*
```{r}
# Weekday vs Weekend trends
weekday_weekend_trends <- data %>%
  group_by(day_type, member_casual) %>%
  summarise(
    mean_trip_duration = mean(trip_duration, na.rm = TRUE),
    trip_count = n()
  )
print(weekday_weekend_trends)

# Plot: Weekday vs Weekend Trends in Trip Duration
ggplot(weekday_weekend_trends, aes(x = day_type, y = mean_trip_duration, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Weekday vs Weekend Trends in Trip Duration") +
  xlab("Day Type") +
  ylab("Average Trip Duration (minutes)") +
  theme_minimal()

```
*Compare trip durations during peak hours (morning/evening) vs. off-peak hours.*
```{r}
# Peak hours (7-10 AM, 4-7 PM) vs off-peak
data$peak_offpeak <- ifelse(data$hour %in% c("07", "08", "09", "10", "16", "17", "18", "19"), "Peak", "Off-Peak")
peak_offpeak_trends <- data %>%
  group_by(peak_offpeak, member_casual) %>%
  summarise(
    mean_trip_duration = mean(trip_duration, na.rm = TRUE),
    trip_count = n()
  )
print(peak_offpeak_trends)

# Plot: Peak vs Off-Peak Trip Trends by Rider Type
ggplot(peak_offpeak_trends, aes(x = peak_offpeak, y = mean_trip_duration, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Peak vs Off-Peak Trends in Trip Duration by Rider Type") +
  xlab("Time Period") +
  ylab("Average Trip Duration (minutes)") +
  theme_minimal()
```

# Proximity Trends → Close vs. Far Trip Counts
*Compare start station proximity for close vs far trip counts*
```{r}
# Summarize trip counts by proximity
proximity_trends <- data %>%
  group_by(start_station_proximity) %>%
  summarise(trip_count = n())
print(proximity_trends)
```

# MBTA Line Trends → Trip Counts
*Compare different MBTA Lines usage through trip count*
```{r}
# Summarize trip counts by MBTA line
mbta_line_trends <- data %>%
  group_by(start_mbta_line) %>%
  summarise(trip_count = n())
print(mbta_line_trends)
```

## RESEARCH QUESTION BASED EDA (used for presentation)

# Question 1: How does the activity at Blue Bike stations far from MBTA stops differ from stations that are close?
```{r}
# Summarize trip count by proximity
proximity_activity <- data %>%
  group_by(start_station_proximity) %>%
  summarise(
    trip_count = n(),
    mean_trip_duration = mean(trip_duration, na.rm = TRUE),
    median_trip_duration = median(trip_duration, na.rm = TRUE)
  )
print(proximity_activity)

# Plot: Trip Count and Duration by Proximity
ggplot(proximity_activity, aes(x = start_station_proximity)) +
  geom_bar(aes(y = trip_count, fill = start_station_proximity), stat = "identity") +
  ggtitle("Activity at Blue Bike Stations by Proximity to MBTA Stops") +
  xlab("Proximity to MBTA") +
  ylab("Trip Count") +
  theme_minimal()
```

```{r}
# Count the number of Blue Bike stations by proximity
stations_proximity <- data %>%
  group_by(start_station_proximity) %>%
  summarise(station_count = n_distinct(start_station_name))
print(stations_proximity)

# Plot: Blue Bike Stations by Proximity to MBTA Stops
ggplot(stations_proximity, aes(x = start_station_proximity, y = station_count, fill = start_station_proximity)) +
  geom_bar(stat = "identity") +
  ggtitle("Blue Bike Stations by Proximity to MBTA Stops") +
  xlab("Proximity to MBTA") +
  ylab("Number of Blue Bike Stations") +
  theme_minimal() +
  theme(legend.position = "none")
```
# Question 2: What MBTA lines/stops are most popular with Blue Bike stations?
```{r}
# Count trips by MBTA line
mbta_line_popularity <- data %>%
  group_by(start_mbta_line) %>%
  summarise(
    trip_count = n()
  ) %>%
  arrange(desc(trip_count))
print(mbta_line_popularity)

# Plot: Popularity of MBTA Lines
ggplot(mbta_line_popularity, aes(x = reorder(start_mbta_line, -trip_count), y = trip_count, fill = start_mbta_line)) +
  geom_bar(stat = "identity") +
  ggtitle("Popularity of MBTA Lines with Blue Bike Stations") +
  xlab("MBTA Line") +
  ylab("Trip Count") +
  scale_fill_manual(values = mbta_colors) + # Apply custom colors
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
# Question 3: Does Blue Bike usage align with the expected travel commute times of MBTA users (7-9am and 5-7pm during the week?
```{r}
# Extract day of the week and hour from the dataset
data$day_of_week <- weekdays(as.Date(data$started_at))
data$hour <- as.numeric(format(as.POSIXct(data$started_at), "%H"))


# Categorize time into peak commute and non-peak
data$commute_period <- ifelse(data$hour %in% c(7, 8, 9, 17, 18, 19), "Commute", "Non-Commute")

# Analyze usage during commute vs. non-commute times
commute_analysis <- data %>%
  group_by(commute_period, member_casual) %>%
  summarise(
    trip_count = n(),
    mean_trip_duration = mean(trip_duration, na.rm = TRUE)
  )
print(commute_analysis)

# Plot: Commute vs Non-Commute Usage Rider Type Comparison (Bar Plot)
ggplot(data, aes(x = commute_period, fill = member_casual)) +
  geom_bar(position = "dodge") +
  ggtitle("Blue Bike Usage During Commute vs Non-Commute Times") +
  xlab("Time Period") +
  ylab("Trip Count") +
  theme_minimal()

# Ensure day_of_week is in chronological order
data$day_of_week <- factor(data$day_of_week, 
                           levels = c("Monday", "Tuesday", "Wednesday", "Thursday", 
                                      "Friday", "Saturday", "Sunday"))

# Plot: Daily Usage Rider Type Comparison (Bar Plot)
ggplot(data, aes(x = day_of_week, fill = member_casual)) +
  geom_bar(position = "dodge", stat = "count") +
  ggtitle("Daily Blue Bike Usage by Rider Type") +
  xlab("Day of Week") +
  ylab("Number of Trips") +
  theme_minimal()

```
# Additionally, does member usage of Blue Bikes decrease on the weekend?
```{r}
# Categorize days into weekday vs weekend
data$day_type <- ifelse(data$day_of_week %in% c("Saturday", "Sunday"), "Weekend", "Weekday")

# Analyze member and casual usage by day type
member_usage_weekend <- data %>%
  group_by(day_type, member_casual) %>%
  summarise(
    trip_count = n(),
    mean_trip_duration = mean(trip_duration, na.rm = TRUE)
  )
print(member_usage_weekend)

# Plot: Member and Casual Usage by Day Type
ggplot(member_usage_weekend, aes(x = day_type, y = trip_count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Blue Bike Usage by Day Type (Weekday vs Weekend)") +
  xlab("Day Type") +
  ylab("Trip Count") +
  theme_minimal()

# Plot: Blue Bike Usage by Hour and Day
usage_heatmap <- data %>%
  group_by(day_of_week, hour) %>%
  summarise(trip_count = n())

ggplot(usage_heatmap, aes(x = hour, y = day_of_week, fill = trip_count)) +
  geom_tile() +
  ggtitle("Heatmap of Blue Bike Usage by Hour and Day") +
  xlab("Hour of Day") +
  ylab("Day of Week") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  theme_minimal()
```